A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-03-26 paper-hierarchical.tex
working...

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performances. Parameters such as flux, beam size at the image and coherence properties are obtained from simple hand-calculations to sophisticated computer simulations using ray tracing and waveoptics techniques. A last generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues 
related to highly demagnifying sychrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storages rings: the old ESRF (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux, intensity distribution and focal size, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS. They are discussed with emphasis in benefits and limitations of each one. 
\end{abstract}

@@

4) OASYS

The paper is structures as follows: we first introduce in Section~\ref{Bealine description} the main parameters of the beamline and make the selection of the particular choice of the parameters used in the calculations. Section~\label{Simulations} is the central part of the paper, and decribe the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In subsection \ref{hand} we calculate with simple methods approximated values of focal size, flux and coherent fraction. Then accuracy is improved using geometrical methods in subsection \ref{ray tracing} that give reasonable results as far as some corrections are considered because of the  mentioned characteristics of high demagnification and high coherence of the studied beamline. The wave optics...

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF. This beamline was built during the Upgrade Phase I at ESRF in \inred{2011??}. It uses one or several undulators as source. During the period 2011-2018 it worked exploiting the ESRF storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. The replacement of this lattice by a new hybrid multi-bend achromat (HMBA) lattice is being achieved in the ESRF Upgrade Phase II. This new lattice will drastically improve the performances of the ESRF source. The result, the Extreme Brilliant Source (EBS) \cite{ESRF2014} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circunference and electron energy $E_e$. As as result, the x-ray sources will be much coherent, specially in the horizontal direction. How this important improvement will afeect a particular beamline is analyzed in this paper.

@@

\begin{table}\label{tab:TableSources}
    \caption{ESRF and ESRF-EBS main storage ring parameters. Main parameters of the present DBA lattice (ESRF) and futurue HMBA lattice (EBS) storage rings \cite{ESRF2014}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (Table \ref{tab:Undulators}). This section is devided in three lain subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

We first estimate the beam size at different positions at the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be done by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between imaging element and image. The demagnification is defined, then, as the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifyng. The latter will be our case, thus is mor comfortable to work with demagnification (>1) rather than magnification.

The source size is usually approximated as the convolution of the electron sizes (Table \ref{tab:eBeam}) with undulator natural source size (zero emittance) \cite{elleaume}. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the electron beam size and divergence at a centre of the symmetry of the straight section. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively. The undulator natural source size and divergence are given by:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u\approx\frac{2.74}{4\pi}\sqrt{\lambda L} && \sigma_u'\approx0.69\sqrt{\frac{\lambda}{L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\lambda_u$ is the undulator length, with $\lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are given by:
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

@@

The focal sizes were estimated, in a first approximation, by multiplying the photon source size by the magnification factors. For predicting the photon beam footprint downstream the beamline and to be able to estimate the needed dimensions of the optical elements to accept the whole beam, it is important to be able to calculate the beam divergence. In an ideal {\'{e}}tendue preserving optical system, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}), one obtains the photon source sizes and divergences shown in Table~\ref{tab:HandCalculations}. The focal size would be 139$\times$5 nm$^2$ (H$\times$V) for ESRF and 10$\times$5 nm$^2$ for EBS. The focal spot for ESRF is too large in horizontal for the demandes beamline specification, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8 \inred{$\approx$ 403.8/50=8.076}. Therefore the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

@@

The next step would be to estimate the number of photons at the sample. For that, we have to know the spectrum at the source. It is not straightforward to get this information ``by hand'' but one can calculate the flux and power using different codes. Perhaps the most performant and well maintained codes spread in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. In this paper, we use the OASYS suite \cite{codeOASYS} that implements in its XOPPY add-on three codes: US\cite{codeUS}, URGENT: \cite{codeUS} and SRW \cite{codeSRW}. Figure~\ref{fig:FluxU18} shows the spectra for two slit apertures calculated with SRW. The spectrum is calculated over an aperture opened to fully accept the central cone at the resonance peak of the harmonic in use. placed at d=27.1. An aperture size of $S_h=S_v=4$mm  because it is much larger than the Full Width at Half Maximum (FWHM) of the central cone at $d$, that is $\max(\Sigma') d$, with any possible value of $\Sigma'$ in Table~\ref{tab:HandCalculations}. The peak intensity is about  $3.5\cdot10^{14}$ photons/s/0.1{\%}bw (see \ref{fig:FluxU18}). The corresponding spectral peak is very wide and depends mostly on the aperture so it does not change when reducing emittance (from high-$\beta$ (ESRF) to EBS). For smaller apertures (1~mm) the emittance plays an important role in both peak width and peak amplitude. The number of photons at the source at photon energy of 17225 eV \inred{(3.5 $10^{14}$ /17.225 photons/s/eV)} has to be multiplied by the bandpass of the monochromator at that energy \inred{(17225 $10^{-2}$)}, thus resulting $N_{ML}$=3.5 $10^{15}$ photons/s.

@@

Now we have to consider which optical element crop the beam removing photons. This can be analyzed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50 \mu$m in horizontal and fully opened in vertical. For ESRF its transmittivity is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their length to a direction perpendicular to the optical axis. Using mirror length of 60 mm and 26 mm for VFM and HFM, respectively, and incident angle of 15 mrad mrad, the KB is behaving like an aperture of $900 \times 390 \mu m^2$. Dividing this by the distance from the mirror to the VSS in H and to the source in V we get N.A. of6.2$\times$21.6 $mu$rad. Dividing by the beam divergence we have the new transmittivity coefficients 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 7.53 $10^{12}$ photons for ESRF and 1.03 $10^{14}$ photons for EBS. \todo{check with referente 6 $10^9$ at 33.6 keV}

>>>>

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performances. Parameters such as flux, beam size at the image and coherence properties are obtained from simple hand-calculations to sophisticated computer simulations using ray tracing and waveoptics techniques. A last generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues 
related to highly demagnifying sychrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storages rings: the old ESRF (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux, intensity distribution and focal size, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis in benefits and limitations of each one. 
\end{abstract}

@@

X-ray beamlines are very particular optical systems. Some implications of exploiting low wavelengths such as the necessity of using grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators make useless, or at least incomplete, the application of common commercial software packages for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), for mentioning only codes used in this work. These historical tools together with new tools designed ad-hoc packages have been agglutinated in a user friendly software OASYS. Many of the calculations presented here have been performed directly in OASYS or with script created with the help of OASYS.

The paper is structures as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and make the selection of the particular choice of the parameters used in the calculations. Section~\label{Simulations} is the central part of the paper, and describe the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In subsection \ref{level0} we calculate with simple methods approximated values of focal size, flux and coherent fraction. Then accuracy is improved using geometrical methods in subsection \ref{level1} that give reasonable results as far as some corrections are considered because of the  mentioned characteristics of high demagnification and high coherence of the studied beamline. The wave optics section~]ref{level2} also include different models whose complexity increases when the results expected are more and mode demanding. The simulations methods are presented in a hierarchical approach:  they provide more and better results but the price to pay is to deal with more complex and computationally expensive tools.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF in \inred{2011??}. It uses one or several undulators as source. During the period 2011-2018 it worked exploiting the ESRF storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. The replacement of this lattice by a new hybrid multi-bend achromat (HMBA) lattice is being achieved in the ESRF Upgrade Phase II. This new lattice will drastically improve the performances of the ESRF source. The result, the Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circunference and electron energy $E_e$. As as result, the x-ray sources will be much coherent, specially in the horizontal direction. This work shows some quantitative results on how the storage ring upgrade will affect a particular beamline is analyzed in this paper.

@@

\begin{table}\label{tab:TableSources}
    \caption{ESRF and ESRF-EBS main storage ring parameters. Main parameters of the present DBA lattice (ESRF) and future HMBA lattice (EBS) storage rings \cite{orangebook}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (Table \ref{tab:Undulators}). This section is devided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

We first estimate the beam size at different positions at the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be done by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between imaging element and image. The demagnification is the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifying. The latter will be our case, thus is more convenient to work with demagnification ($M^{-1}>1$) rather than with magnification.

The source size is usually approximated supposing a Gaussian distribution of the electron sizes (or divergence) (Table \ref{tab:eBeam}) and undulator natural source size or divergence. Although the former is true, the latter is not. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the electron beam size and divergence at a centre of the symmetry of the straight section. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively. The undulator natural source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \sqrt{\frac{\lambda L}{2 \pi^2}}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\lambda_u$ is the undulator length, with $\lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are given by the convolution of the electron sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

@@

The beam sizes at different positions at the beamline were estimated in a different way depending if the beam is focused or not. If the beam is focused, its dimensions in a first approximation are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam in a non-focal position the beam divergence is used. In an ideal {\'{e}}tendue preserving optical system, where the conservation of the Smith-Helmholz invariante applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}), one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The focal spot for ESRF lattice is too large in horizontal for the beamline specification, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8 \inred{$\approx$ 403.8/50=8.076}. Therefore the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

@@

The next step would be to estimate the number of photons at the sample. For that, we need the spectrum at the source. It is not straightforward to get this information ``by hand'' but one can calculate the flux and power using different codes. Perhaps the most performant and well maintained codes spread in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US\cite{codeUS}, URGENT: \cite{codeUS} and SRW \cite{codeSRW}. Figure~\ref{fig:FluxU18} shows the spectra for two slit apertures calculated with SRW. The spectrum is calculated over an aperture opened to fully accept the central cone at the resonance peak of the harmonic in use. placed at $d$=27.1m. An aperture size of $S_h=S_v=4$ mm it is much larger than the Full Width at Half Maximum (FWHM) of the central cone at $d$, that is $\max(\Sigma') d$ \todo{HOW MUCH?}, with any possible value of $\Sigma'$ in Table~\ref{tab:HandCalculations}. The peak intensity is about  $3.5\cdot10^{14}$ photons/s/0.1{\%}bw (see \ref{fig:FluxU18}). The corresponding spectral peak is very wide and depends mostly on the aperture so it does not change when reducing emittance (from high-$\beta$ (ESRF) to EBS). For smaller apertures (1~mm) the emittance plays an important role in both peak width and peak amplitude. The number of photons at the source at photon energy of 17225 eV \inred{(3.5 $10^{14}$ /17.225 photons/s/eV)} has to be multiplied by the bandpass of the monochromator at that energy \inred{(17225 $10^{-2}$)}, thus resulting $N_{ML}$=3.5 $10^{15}$ photons/s.

@@

\inred{Next we would like to estimate} the geometrical transmitivity of the beamline, therefore we have to consider which optical element crop the beam and remove photons. This can be analysed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50 \mu$m in horizontal and fully opened in vertical. For ESRF its transmittivity is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their length to a direction perpendicular to the optical axis. Using mirror length of 60 mm and 26 mm for VFM and HFM, respectively, and incident angle of 15 mrad mrad, the KB is behaving like an aperture of $900 \times 390 \mu m^2$. Dividing this by the distance from the mirror to the VSS in H and to the source in V we get N.A. of6.2$\times$21.6 $mu$rad. Dividing by the beam divergence we have the new transmittivity coefficients 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 7.53 $10^{12}$ photons for ESRF and 1.03 $10^{14}$ photons for EBS. \todo{check with referente 6 $10^9$ at 33.6 keV}
